On the presentation of orthogonal groups over polynomial rings

### 2019-07-03 dl-paper.tex
Rewritten Lemma 3.3

For $n\geq 1$ we define ``truncated'' Steinberg group $\St^{\leq n}(\Phi, B)$ as the quotient of the free group generated by the set $\mathcal{X}_{\leq n}^\Phi$ of homogeneous Steinberg generators of degree $\leq n$ modulo the normal subgroup $\mathcal{R}_{\leq n}^\Phi$ generated by the set of Steinberg relations of degree $\leq n$.

@@

\begin{prop} For $\Phi=\rA_{\geq 4}, \rD_{\geq 5}, \rE_{6,7,8}$ and $n \geq 1$ the map $i_n\colon \St^{\leq n}(\Phi, B) \to \St^{\leq n+1}(\Phi, B)$ induced by the natural embedding of generators, is an isomorphism. In particular, the obvious map $\St^{\leq 1}(\Phi, B) \to \St(\Phi, B)$ is an isomorphism. \end{prop}
\begin{proof}
 The special case $\Phi=\rA_{\geq 4}$ of the assertion has been proved by Tulenbayev (see~\cite[Lemma~3.3]{Tu83}) 
 and therefore can be used to deduce the assertion in the cases $\Phi=\rD_\ell,\rE_\ell$.

 %Denote by $A_4(\Phi)$ the set consisting of all root subsystems of $\Phi$ of type $\rA_4$ and by $G_n$ the free product of groups $\coprod\limits_{\Psi \in A_4(\Phi)} \St^{\leq n}(\Psi, B)$ amalgamated over relations $x^{\Psi_1}_\alpha(a) = x^{\Psi_2}_\alpha(a)$ for all $a \in B_d$, $d \leq n$, $\alpha \in \Psi_1\cap\Psi_2$, $\Psi_1, \Psi_2 \in A_4(\Phi)$.

 Let us show that the obvious map $i_n \colon G_n \to G_{n+1}$ has an inverse $j_n$. Notice first that the maps $j_\Psi$ constructed by Tulenbaev agree on the generators of $G_{n+1}$ and hence, by the universal property of colimits, give rise to the well-defined map $j\colon G_{n+1}\to G_n$, which is easily seen to be  inverse to $i$.
  \begin{equation} \nonumber \xymatrix{
 \St^{\leq n+1}(\Psi, B) \ar[r]^{j_\Psi} \ar[d] & \St^{\leq n}(\Psi, B) \ar[d] \\
 G_{n+1} \ar[r]^{j_n} & G_{n} }
 \end{equation}
Notice that in the case $\Phi = \rE_\ell$ the natural homomorphisms $G_k \to \St^{\leq k}(\rE_\ell, B)$ are isomorphisms for $k=n,n+1$ (cf.\cite[Corollary 10]{S15}), therefore the proof is already complete in this case.

We are left to consider the case $\Phi = \rD_\ell$. Notice that any pair of nonperpendicular roots  $\alpha, \beta$ is contained in some subsystem $\Psi \subseteq \Phi$ of type $\rA_4$, hence the only defining relations of $\St^{\leq n+1}(\Phi, B)$ which do not occur already in $G_{n+1},$ are $\text{R3}^\bot_{d, n+1}$ for $d \leq n+1$ (more precisely, the relations for which $\alpha$ and $\gamma$ are orthogonal and lie in a root subsystem of type $\rD_3$, cf.~\cite[Remark~2.1]{S15}).
\end{proof}

>>>>

For $n\geq 1$ we define ``truncated'' Steinberg group $\St^{\leq n}(\Phi, B)$ by means of the set $\mathcal{X}_{\leq n}^\Phi$ of homogeneous Steinberg generators of degree $\leq n$ and the set of Steinberg relations $\mathcal{R}_{\leq n}^\Phi$ of degree $\leq n$.

@@

\begin{prop} For $\Phi=\rA_{\geq 4}, \rD_{\geq 5}, \rE_{6,7,8}$ and $n \geq 1$ the map $i_n\colon \St^{\leq n}(\Phi, B) \to \St^{\leq n+1}(\Phi, B)$ induced by the natural embedding of generators, is an isomorphism. In particular, the obvious map $\St^{\leq 1}(\Phi, B) \to \St(\Phi, B)$ is an isomorphism. \end{prop}
\begin{proof}
 We need to construct a map $j_n$ which would be the inverse of $i_n$. 
 We start with a word map $\widetilde{j}_n^\Phi \colon F\langle \mathcal{X}^\Phi_{\leq n+1} \rangle \to \St^{\leq n}(\Phi, B)$ defined via
 \[ \widetilde{j}_n(x_\alpha(at^k)) = \begin{cases} x_\alpha(at^k), & k\leq n; \\
      [x_{\alpha - \beta} (N_{\alpha-\beta, \beta} \cdot at^{k-1}), x_{\beta}(t)], & k = n+1, \end{cases} \]
 where $\beta$ is an arbitrary root of $\Phi$ forming a sharp angle with $\alpha$.
 Standard argument (cf.~\cite[Proposition~1.1]{Re75} or~\cite[Proposition~3.2.2]{RS76}) shows that $\widetilde{j}_n$ does not depend on the choice of $\beta$.

 Set $\mathcal{R}^\Phi_{n+1} = \mathcal{R}^\Phi_{\leq n+1} \setminus \mathcal{R}^\Phi_{\leq n}$.
 We need to verify that the image of every relator $R \in \mathcal{R}^\Phi_{n+1}$ under $\widetilde{j}^\Phi_n$ is a trivial element of $\St^{\leq n}(\Phi, B)$. In the special case $\Phi=\rA_{\geq 4}$ this assertion has already been demonstrated by Tulenbayev in~\cite[Lemma~3.3]{Tu83}. We will deduce the assertion in the remaining cases $\Phi=\rD_\ell,\rE_\ell$ from this Tulebayev's result.

 Fix a relator $R \in \mathcal{R}^\Phi_{n+1}$. By~\cref{superfluous-relations} we may assume that $R$ is not of type $\text{R3}^\bot$, hence the roots $\alpha, \beta$ appearing in $R$ are contained in a root subsystem of $\Phi$ of type $\rA_2$. Our assumptions on $\Phi$ guarantee that there exists some root subsystem $\Psi$ of type $\rA_4$ containing $\alpha$ and $\beta$. Consider the following commutative diagram in which the vertical arrows are induced by the embedding $\Psi\subseteq\Phi$. 
  \begin{equation} \nonumber \xymatrix{
 F(\mathcal{X}_{\leq n+1}^\Psi) \ar[r]^{j_n^\Psi} \ar[d] & \St^{\leq n}(\Psi, B) \ar[d] \\
 F(\mathcal{X}_{\leq n+1}^\Phi) \ar[r]^{j_n^\Phi} & \St^{\leq n}(\Phi, B) }
 \end{equation}
The relation $R$ lies in the image of the left arrow, therefore it comes from some relation $R' \in \mathcal{R}^\Psi_{n+1}$. The image of $R'$ in $\St^{\leq n}(\Psi, B)$ under $j_n^\Psi$ is trivial by Tulenbayev's result. But this implies that the image of $R$ under $\widetilde{j}_n^\Phi$ is also trivial and hence that $\widetilde{j}_n^\Phi$ gives rise to the desired map $j_n$.
\end{proof}
